
---
title: "Treasury Analytical Note : Insights from New Zealand child poverty data"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(data.table)
library(ggplot2)
library(scales)
library(knitr)
library(gt)
library(dplyr)
library(patchwork)
library(gt)
source("src/utilities.R")

###########################
## read in data
###########################

filename <-  "data/CPClustering_final.csv"
groupname <- "Characteristics_cluster_3"
ordering <- fread("data/order.csv")
extra_results_file <- "data/results_for_release_AdditionalMW.csv"

###########################
# rename and group
###########################

name_map <- c("0" = "Not near poverty", 
              "3" = "Ben1",   "1" = "Ben2", "2" = "Ben3",
              "5" = "Working1", "7" = "Working2", "6" = "Working3", "4" = "Working4")

exclude_group <- "Not near poverty"

cluster_characteristics <- c(
  "H_BHC_Decile" = "BHC (equivalised household income, before housing costs) decile",
  "H_HCost_Prop_Decile" = "Housing costs as a proportion of household income",
  "H_Dep17" = "Dep17", 
  "F_Income_Source" = "Family's main income source"
)

poverty_characteristics <- c("In_BHC_Poverty" = "Below moving-line BHC50 threshold", 
                             "In_Fixed_AHC_Poverty" = "Below fixed-line AHC50 threshold", 
                             "In_MH_Poverty" = "Material Hardship (Dep17>=6)", 
                             "Poverty" = "Poverty")

 
main_characteristics <- c("H_Dep17" = "Dep17", 
                          "AHC_Depth_Band" = "Dollars from fixed AHC50 threshold", 
                          "BHC_Depth_Band" = "Dollars from relative BHC50 threshold", 
                          "Poverty" = "Poverty indicators",
                          "F_Income_Source" = "Family's main income source", 
                          "H_HCost_Prop_Decile" = "Housing costs as a proportion of household income", 
                          "H_Tenure_Detailed" = "Housing tenure", 
                          "F_Is_ASAdmin" = "Family is estimated to receive Accommodation Supplement",    
                          "F_Eligible_AS_NotRecieving" = "Family is modelled to be eligible for Accommodation Supplement but is not receiving it", 
                          "F_Disabled" = "Someone in the family has a disability",
                          "F_ParentDisabled" = "A parent in the family has a disability", 
                          "F_Highest_Qual" = "Parents' highest qualification", 
                          "F_Is_Couple" = "Family type", 
                          "H_Crowded" = "Is the house crowded?",
                          "F_N_Earners" = "Number of earners in the family", 
                          "H_People" = "Number of people in the household")

additional_characteristics <- 
  c(
    "H_AHC_Decile" = "AHC (equivalised household income, after housing costs) decile",
    "H_BHC_Decile" = "BHC (equivalised household income, before housing costs) decile",
    "AHC_Depth_Band" = "Dollars from fixed AHC50 threshold", 
    "BHC_Depth_Band" = "Dollars from relative BHC50 threshold",
    "Disabled" = "Disabled child", 
    "F_ChildDisabled" = "Disabled child in the family", 
    "F_Disabled" = "Disabled person in the family", 
    "F_ParentDisabled" = "Disabled parent",
    "F_Is_Beneficiary" = "Family received a core benefit at some time over the year",
    "F_Eligible_AS" = "Family is modelled to be eligible for Accommodation Supplement", 
    "F_Eligible_AS_NotRecieving" = "Family is modelled to be eligible for Accommodation Supplement but is not receiving it", 
    "F_Is_WFF" = "Family is modelled to be eligible for Working for Families",
    "H_ASArea" = "Accommodation Supplement area", 
    "H_Crowded" = "Is the house crowded",
    "H_Dep17" = "Deprivation Index (decile)", 
    "H_HCost_Prop_Decile" = "Housing costs as a proportion of household income", 
    "H_HCost_Decile" = "Housing cost (decile)",   
    "H_Region" = "Region",
    "H_Tenure_Detailed" = "Tenure", 
    "Asian" = "Asian", 
    "European" = "European",
    "Maori" = "Māori", 
    "Pacific" = "Pacific",
    "F_Is_Couple" = "Family type", # change 
    "F_Num_Children_Class" = "Number of children in the family",
    "Age" = "Age of child", 
    "F_Youngest" = "Youngest child in the family", 
    "F_Oldest" = "Oldest person in the family",
    "H_People" = "Number of people in the household",
    "H_Families" = "Number of families in the household",
    "F_Highest_Qual" = "Parents' highest qualification", 
    "F_Currently_FTEmployed" = "A parent is currently working full time (according to HES)",
    "F_N_Earners" = "Number of earners in the family", 
    "F_Private_Decile" = "Family private (non-benefit) income (decile)",
    "F_Income_Source" = "Family's main income source"
  )

cluster_characteristics <- c(
  "H_BHC_Decile" = "BHC (equivalised household income, before housing costs) decile",
  "H_HCost_Prop_Decile" = "Housing costs as a proportion of household income",
  "H_Dep17" = "Dep17", 
  "F_Income_Source" = "Family's main income source"
)

MW_characteristics <- c("HadToBorrow" = "In the last twelve months, have you borrowed from friends or family to meet everyday living costs because of a shortage of money?", 
                        "BrokenApplicances" = "In the last twelve months have you delayed replacing or repairing broken or damaged appliances to keep costs down?", 
                        "Cold" = "In the last twelve months have you put up with feeling cold to keep costs down?",         
                        "IsYourIncomeEnough" = "Is your income enough?", 
                        "LessMeat" = "In the last twelve months have you bought cheaper cuts of meat or less meat than you would like to keep costs down?", 
                        "LimitedClothesShoes" = "When buying, or thinking about buying, clothes or shoes for yourself do you usually feel limited by money available?", 
                        "NoInsurance" = "Do you have house insurance?", 
                        "NoLocalTrips" = "In the last twelve months have you done without or cut back on trips to the shops or other local placesto keep costs down?", 
                        "NoMeal" = "Do you have a meal with meat, fish or chicken (or vegetarian equivalent) at least each second day?", 
                        "NoShoes" = "Do you have two pairs of shoes in a good condition that are suitable for your daily  activities?", 
                        "NoDentist" = "In the last twelve months have you had to postpone or put off visits to the dentist to keep costs down?", 
                        "NoDoctor" = "In the last twelve months have you had to postpone or put off visits to the doctor to keep costs down?", 
                        "NoFruitVeg" = "In the last twelve months have you had to go without fresh fruit or vegetables to keep costs down?", 
                        "NoGifts" = "Do you give presents to family or friends on birthdays, Christmas or other special occasions?",
                        "NoSuitableClothes" = "Do you have suitable clothes for important or special occasions?", 
                        "NotPaidCar" = "In the last twelve months, have you not been able to pay for car insurance, registration or warrant of fitness on time because of a shortage of money?", 
                        "NotPaidUtility" = "In the last twelve months, have you not been able to pay electricity, gas, rates or water bills  on time because of a shortage of money?", 
                        "UnexpectedExpense" = "If you (or your partner) had an unexpected and unavoidable expense of $500 in the next week, could you pay in a month without borrowing?")

palette_percent <- col_numeric(palette = "viridis", domain = c(0, 1), alpha = 0.75, na.color ="#440154FF")
palette_risk <- col_numeric(palette = "viridis", domain = c(0, 3), alpha = 0.75, na.color =  "#440154FF") 

all_results <- characteristics(filename, 
                               groupname, 
                               name_map, 
                               exclude_group = "Not near poverty")[[2]]

# fix some MW
extra_results <- characteristics(extra_results_file , 
                               "Characteristics_cluster_3", 
                               name_map, 
                               exclude_group = "Not near poverty")[[2]]
all_results <- all_results[!(Descriptor_1 %in% extra_results[, Descriptor_1])]
all_results <- rbind(all_results, 
                     extra_results)

# better names
all_results[Descriptor_1=="F_Is_Couple" & Descriptor_2 == FALSE, Descriptor_2 := "Single parent"]
all_results[Descriptor_1=="F_Is_Couple" & Descriptor_2 == TRUE, Descriptor_2 := "Coupled parent"]

all_results[Descriptor_1 %in% names(MW_characteristics) & Descriptor_2 == FALSE, Descriptor_2 := "No Deprivation"]
all_results[Descriptor_1 %in% names(MW_characteristics) & Descriptor_2 == TRUE, Descriptor_2 := "Deprivation"]

HCost_prop_names <- c("1" = "0%", "2" = "3%-6%", "3" = "6%-10%", 
                      "4" = "10%-15%", "5" = "15%-20%", "6" = "20%-23%", 
                      "7" = "23%-28%", "8" = "28%-35%", "9" = "35%-46%", 
                      "10" = ">46%" )
all_results[Descriptor_1 == "H_HCost_Prop_Decile" , Descriptor_2 := HCost_prop_names[Descriptor_2]]

FPrivateIncome_names <- c("1" = "<$1", "2" = "$1-$3,500", "3" = "$3,500-$15,000", 
                      "4" = "$15,000-$32,000", "5" = "$32,000-$48,000", "6" = "$48,000-$65,000", 
                      "7" = "$65,000-$88,000", "8" = "$88,000-$118,000", "9" = "$118,000-$162,000", 
                      "10" = ">$162,000" )
all_results[Descriptor_1 == "F_Private_Decile" , Descriptor_2 := FPrivateIncome_names[Descriptor_2]]



```


These detailed outputs accompany the report https://www.treasury.govt.nz/publications/an/an-22-04

This analytical note outlines an approach that uses the available data to provide insights into three different indicators of poverty, making use of recent data and modelling improvements. It applies a statistical algorithm to identify seven different categories of children in poverty and describes the characteristics of children in each group.

To focus on children near or under at least one poverty threshold, we defined our population of interest to be in households with either equivalised before-housing-cost incomes in the bottom 20%, after-housing-cost incomes in the bottom 20%, or Dep17 scores of 5 or more. This includes the 270,000 children who are in poverty according to at least one of the main indicators but is a larger group including a total of 360,000 children (approximately 30% of children in New Zealand). 

The clustering algorithm described in the report identified seven categories within this population. Children in families who are mainly supported by core benefits represent three groups, and children in families that are mainly supported by market income represent the remaining four. Some children could be considered to fall between groups, so the total numbers of children in each group should be considered indicative. 

The tables below describe the characteristics of each group. 


## Groups

* Ben1: Beneficiary, extreme hardship

* Ben2: Beneficiary, hardship with high housing costs

* Ben3: Beneficiary, low income with some hardship

* Working1: Working, extreme hardship

* Working2: Working, hardship

* Working3: Working, high housing costs but no hardship  

* Working4: Working, low incomes but no hardship

These detailed results have been derived by the author using the TAWA model. They are indicative of the characteristics of each group. 

## Modelling assumptions

* This analysis was carried out using Treasury’s micro-simulation model of the tax and welfare system - TAWA. All calculations should be considered as estimations.

* Estimates are for the 2019/20 tax year using HES 2019/20 augmented using IDI data.

* The modified OECD equivalisation scale is used to calculate equivalised incomes.

* A take-up rate (a model based on admin recipients) has been applied to the Accommodation Supplement when the expected number of recipients changes. Recipients are in the first case based on receipt in the admin data.

* Apparent inconsistencies in totals are due to rounding and/or suppression, with estimates being suppressed if they did not meet the confidentiality requirements of Statistics NZ.

## IDI Disclaimer

These results are not official statistics. They have been created for research purposes from the Integrated Data Infrastructure (IDI) which is carefully managed by Stats NZ. For more information about the IDI please visit https://www.stats.govt.nz/integrated-data/. The results are based in part on tax data supplied by Inland Revenue to Stats NZ under the Tax Administration Act 1994 for statistical purposes. Any discussion of data limitations or weaknesses is in the context of using the IDI for statistical purposes, and is not related to the data’s ability to support Inland Revenue’s core operational requirements.



## Clustering characteristics {.tabset .tabset-pills .tabcontent}

### Number of Children

```{r  echo=FALSE}
make_table(all_results, cluster_characteristics, "Population")

```

### Item Probability

Item probability: what proportion of children in this group have this characteristic

```{r echo=FALSE}
make_table(all_results, cluster_characteristics, "Item_Prob")
```


### Relative Risk (compared to other groups)
Relative risk (near poverty): compared to the rest of the population of interest, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group
```{r echo=FALSE, warning=FALSE, warning=FALSE}

make_table(all_results, cluster_characteristics, "Relative_Risk_PopOfInterest", exclude_group)
```

### Relative Risk (compared to population)

Relative risk (all): compared to the rest of the population, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group

```{r echo=FALSE, warning=FALSE, warning=FALSE}

make_table(all_results, cluster_characteristics, "Relative_Risk_all")
```

## Poverty {.tabset .tabset-pills .tabcontent}

### Number of Children

```{r  echo=FALSE}
make_table(all_results, poverty_characteristics, "Population")

```

### Item Probability

Item probability: what proportion of children in this group have this characteristic

```{r echo=FALSE}
make_table(all_results, poverty_characteristics, "Item_Prob")
```


### Relative Risk (compared to other groups)
Relative risk (near poverty): compared to the rest of the population of interest, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group
```{r echo=FALSE, warning=FALSE, warning=FALSE}

make_table(all_results, poverty_characteristics, "Relative_Risk_PopOfInterest", exclude_group)
```

### Relative Risk (compared to population)

Relative risk (all): compared to the rest of the population, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group

```{r echo=FALSE, warning=FALSE, warning=FALSE}

make_table(all_results, poverty_characteristics, "Relative_Risk_all")
```


## Main characteristics {.tabset .tabset-pills .tabcontent}

### Number of Children {.tabset .tabset-pills .tabcontent}

```{r, results='asis', echo=FALSE}
for (category in names(main_characteristics)){
  
  cat("#### ", main_characteristics[category], "\n")
  print(make_table(all_results,main_characteristics[category], "Population"))
  cat('\n\n')
}

```


### Item Probability {.tabset .tabset-pills .tabcontent}

Item probability: what proportion of children in this group have this characteristic

```{r, results='asis', echo=FALSE, warning=FALSE}
for (category in names(main_characteristics)){
  
  cat("#### ", main_characteristics[category], "\n")
  print(make_table(all_results,main_characteristics[category], "Item_Prob"))
  cat('\n\n')
}

```

### Relative Risk (compared to other groups) {.tabset .tabset-pills .tabcontent}

Relative risk (near poverty): compared to the rest of the population of interest, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group

```{r, results='asis', echo=FALSE, warning=FALSE}
for (category in names(main_characteristics)){
  
  cat("#### ", main_characteristics[category], "\n")
  print(make_table(all_results,main_characteristics[category], "Relative_Risk_PopOfInterest",exclude_group))
  cat('\n\n')
}

```


### Relative Risk (compared to population) {.tabset .tabset-pills .tabcontent}

Relative risk (all): compared to the rest of the population, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group

```{r, results='asis', echo=FALSE, warning=FALSE}
for (category in names(main_characteristics)){
  
  cat("#### ", main_characteristics[category], "\n")
  print(make_table(all_results,main_characteristics[category], "Relative_Risk_all"))
  cat('\n\n')
}

```


## Additional characteristics {.tabset .tabset-pills .tabcontent}

### Number of Children {.tabset .tabset-pills .tabcontent}

```{r, results='asis', echo=FALSE}
for (category in names(additional_characteristics)){
  
  cat("#### ", additional_characteristics[category], "\n")
  print(make_table(all_results,additional_characteristics[category], "Population"))
  cat('\n\n')
}

```


### Item Probability {.tabset .tabset-pills .tabcontent}

Item probability: what proportion of children in this group have this characteristic

```{r, results='asis', echo=FALSE, warning=FALSE}
for (category in names(additional_characteristics)){
  
  cat("#### ", additional_characteristics[category], "\n")
  print(make_table(all_results,additional_characteristics[category], "Item_Prob"))
  cat('\n\n')
}

```

### Relative Risk (compared to other groups) {.tabset .tabset-pills .tabcontent}

Relative risk (near poverty): compared to the rest of the population of interest, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group

```{r, results='asis', echo=FALSE, warning=FALSE}
for (category in names(additional_characteristics)){
  
  cat("#### ", additional_characteristics[category], "\n")
  print(make_table(all_results,additional_characteristics[category], "Relative_Risk_PopOfInterest", exclude_group))
  cat('\n\n')
}

```


### Relative Risk (compared to population) {.tabset .tabset-pills .tabcontent}

Relative risk (all): compared to the rest of the population, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group

```{r, results='asis', echo=FALSE, warning=FALSE}
for (category in names(additional_characteristics)){
  
  cat("#### ", additional_characteristics[category], "\n")
  print(make_table(all_results,additional_characteristics[category], "Relative_Risk_all"))
  cat('\n\n')
}

```


## Material Wellbeing characteristics {.tabset .tabset-pills .tabcontent}

### Number of Children

```{r  echo=FALSE}
make_table(all_results, MW_characteristics, "Population")

```

### Item Probability

Item probability: what proportion of children in this group have this characteristic

```{r echo=FALSE}
make_table(all_results, MW_characteristics, "Item_Prob")
```


### Relative Risk (compared to other groups)
Relative risk (near poverty): compared to the rest of the population of interest, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group
```{r echo=FALSE, warning=FALSE, warning=FALSE}

make_table(all_results, MW_characteristics, "Relative_Risk_PopOfInterest", exclude_group)
```

### Relative Risk (compared to population)

Relative risk (all): compared to the rest of the population, how prevalent is this characteristic. 1=No difference, <1=Less likely for this group, >1=more likely for this group

```{r echo=FALSE, warning=FALSE, warning=FALSE}

make_table(all_results, MW_characteristics, "Relative_Risk_all")
```
